---
title: "Tool Time Signature Visualization(s)"
execute:
  echo: false
  warning: false
---

## Lateralus

#### Schism

A rendering of time signatures from the song *Schism* by the band **Tool**. Color is applied when a time signature differs from the value immediately preceding it. Each data point represents one bar and all time signatures have been normalized to sixteenths. From this visualization, you can see there is a section of frequent time signature change from bars 55 to 69.


* [Source](https://musescore.com/user/6831511/scores/4151846)
* [Data (.csv)](https://github.com/asantone/quarto/blob/main/tool/time_signatures.csv)
* [Files (.qmd)](https://github.com/asantone/quarto/tree/main/tool) 

---

```{r}
library(tidyverse)
```

```{r}
#read in the data
main<-read_csv("time_signatures.csv")
```

```{r}
ggplot(main, aes(x=bar, y=sixteens))+
  geom_point(aes(color=factor(different)))+
  scale_colour_manual(values=c("#000000", "#00aaff"))+
  theme_void()+
  theme(legend.position="none")
```

---

## Lateralus

#### Lateralus

A rendering of time signatures from the song *Lateralus* by the band **Tool**. Color is applied when a time signature differs from the value immediately preceding it. Each data point represents one bar and all time signatures have been normalized to sixteenths. From this visualization, you can see there is a section of frequent time signature change from bars 84 to 104 and a total of about 43 time signature shifts. 

Notably, there are also two specifically-identified [polyrhythmic](https://en.wikipedia.org/wiki/Polyrhythm) sections on top of 6/8: a section with 5/8 from bars 111 to 130 and a section of 4/8 from bars 181 to 205. Danny Carey is recognized for [his use of complex time signatures](https://www.drumeo.com/beat/danny-carey-genius/). 


* [Source](https://musescore.com/arindam_octaves101/lateralus)
* [Data (.csv)](https://github.com/asantone/quarto/blob/main/tool/time_signatures.csv)
* [Files (.qmd)](https://github.com/asantone/quarto/tree/main/tool) 

---

```{r}
#readxl needed for sheet-specific data import
library(tidyverse)
library(readxl)
```

```{r}
#read in the data
main<-read_excel("time_signatures.xlsx", "lateralus")

#define a specific color for a second data series
color_polyrhythm<-"#ff3300"
```

```{r}
#visualization but the y values are specified in the geom_point calls
ggplot(main, aes(x=bar))+
  geom_point(aes(y = sixteens, color=factor(different)))+
  geom_point(aes(y = poly_sixteens, color=color_polyrhythm))+
  scale_colour_manual(values=c(color_polyrhythm, "#000000", "#00aaff"))+
  theme_void()+
  theme(legend.position="none")
```